Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-15 persistence.tex
Before meeting on Friday mid July

In this section we present the theory of persistence as is used in connection with geometric questions.
A bridge build through filtrations by sublevel sets of real-valued functions.
For a more complete treatment of persistence theory we refer to \cite{Chazal.2016a, MR3408277}.
An example of interest to us is given by volume functionals on moduli spaces of Riemannian manifolds.
For more examples please consult \cite{polterovich2020topological}.

The so called \textit{persistence pipeline} relates geometry, algebra and discrete mathematics as follows:
Given a space $X$ filtered by a function $f$, the application of the $n^{\mathrm{th}}$ degree part of a homology theory or, more generally, a functor from topological to vector spaces, produces a \textit{persistence module}, an algebraic object equipped with a structure theory leading in favorable cases to a powerful discrete invariant called \textit{persistence diagram}.

These invariants are key for applications.
For example, in the next section we will see how they lead to a generalization of Morse inequalities.
Much of their usefulness comes from a remarkable geometric fact known as stability, stating that the set of all persistence diagrams is a metric space and the map from functions on $X$ with the $L^\infty$ norm to it is 1-Lipschitz.

Let us now explain in detail all the terms used above.
Consider a space $X$ and a function $f \colon X \to \R$.
Unless otherwise noted, the functions we consider need not be continuous.
We pass to filtered spaces by considering the \textit{sublevel set filtration $f_{\leq \bullet}$ of $X$ induced by $f$}, which is defined by
\begin{equation*}
f_{\leq t} = f^{-1}(-\infty, t].
\end{equation*}
We say this filtration is \textit{compact} if $X$ is a locally compact Hausdorff space and all sublevel sets are compact.

For the next step in the persistence pipeline, we will need a coherent assignment of a vector space to any topological space.
More precisely, a functor from $\Top$ to $\Vect$, with key examples coming from \emph{homology theories}, which for now are simply assumed to be $\Z$-graded families $\H = (\H_d)_{d \in Z}$ of \emph{homotopy invariant functors}, which are functors assigning the same morphism to homotopic maps.
Particularly important for us are \v{C}ech homology \cite[Section IX-X]{MR0050886} with field coefficients, and vector-space-valued homology theories in the sense of Eilenberg-Steenrod \cite[Section I]{MR0050886}, such as singular homology with field coefficients \cite{Eilenberg.1944}.

By definition, applying to a filtered space $\{X_t\}_{t \in \R}$ a functor from $\Top$ to $\Vect$, yields for every $t \in \R$ a vector space $M_t$ and for any pair $s, t \in \R$ with $s \leq t$ a linear map $M_{s,t} \colon M_s \to M_t$ such that $M_{t,t}$ is the identity, and the composition $M_{s,t} \circ M_{r,s}$ is equal to $M_{r,t}$ for any triple $r \leq s \leq t$.

In other words, we obtain a functor from the real numbers considered as a category via its linear order structure to the category of vector spaces.
Such functors will be called \emph{persistence modules}.
A morphism of persistence modules $\varphi \colon M \to N$ is a natural transformations, i.e., an assignment of a linear map $\varphi_t \colon M_t \to N_t$ for every $t \in \R$ making the diagram
\begin{equation*}
    \begin{tikzcd}
    M_{s} \arrow[r, "M_{s,t}"] \arrow[d] & M_{t} \arrow[d] \\
    N_{s} \arrow[r, "N_{s,t}"] & N_{t}
    \end{tikzcd}
\end{equation*}
commute for all pairs $s \leq t$.

Usual constructions that work in the category of vector spaces can be transported to the category of persistence modules by performing them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for an interval $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
    C(I)_t =
    \begin{cases}
        \mathbb{F} & \text{if } t \in I, \\
        0          & \text{otherwise},
    \end{cases}
    \qquad
    \qquad
    C(I)_{s, t} =
    \begin{cases}
        \operatorname{id}_{\mathbb{F}} & \text{if } s, t \in I,\\
        0 & \text{otherwise}.
    \end{cases}    
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of them.
% \[
% \bigoplus_{\lambda \in \Lambda} C(I_{\lambda}).
% \]
The multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ associated to a barcode module is known as its \textit{barcode}. By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$. Thus, if a persistence module $M$ is isomorphic to a barcode module, the associated barcode is a complete invariant of $M$, and the (non-unique) isomorphism to the barcode module is called a \emph{barcode decomposition}.

We want to use the passage from persistence modules to invariants obtained from decompositions as the last step in our pipeline, so we have to study which persistence modules admit barcode decompositions.
The most commonly used existence result is Crawley-Boevey's Theorem \cite{Crawley-Boevey.2015}, stating that every persistence module $M$ for which $M_t$ is a finite dimensional vector space for all $t \in \R$ admits a barcode decomposition.
Such persistence modules are called \emph{pointwise finite dimensional ($\PFD$)}.

Unfortunately, the $\PFD$ condition is too restrictive for our purposes. In particular, it is not necessarily satisfied in the historical setup of Morse's work on minimal surfaces. In this context, the persistence modules $M$ only have the slightly weaker property of being \emph{q-tame}, which is defined by the rank of the maps $M_{s,t} \colon M_s \to M_t$ being finite for all $s < t$.

Not every q-tame persistence module admits a barcode decomposition in the above sense, as exemplified by the infinite product $\prod_{n \in \N_{> 0}} C([0,1/n))$, but there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}. We briefly recall the approach from \citet{Chazal.2016b}.

A persistence module $K$ is called \emph{ephemeral} if the maps $K_{s,t} \colon K_s \to K_t$ are $0$ for all $s < t$.
The \emph{radical} $\rad M$ of a persistence module $M$ is defined as the unique maximal submodule of $M$ such that the cokernel of the inclusion $\rad M \to M$ is an ephemeral persistence module.
If $M$ is q-tame, then its radical admits a barcode decomposition \cite[Theorem ?]{Chazal.2016b},
with the associated barcode describing the isomorphism type of $M$ ``up-to-ephemerals".

This can be formalized by constructing the \emph{observable category of persistence modules}, which is equivalent to the quotient of the category of persistence modules by the full sub-category of ephemeral persistence modules.
Intuitively, one may think of the observable category as forgetting all information in persistence modules that does not persist over a non-zero amount of time.
The barcode of the radical of a q-tame persistence module $M$ is then a complete invariant of $M$ in the observable category.

@@

In total, we have seen how to obtain a persistence diagram from a real-valued function $f$ by applying a functor $\H \colon \Top \to \Vect$ to its sublevel set filtration and considering the persistence diagram of the resulting persistence module $\H(f_{\leq \bullet})$, which is well-defined provided that $\H(f_{\leq \bullet})$ is q-tame. If this is the case, we will call the function \emph{q-tame} with respect to $\H$. If $\H$ is some kind of homology, we will call $\H(f_{\leq \bullet})$ the \emph{persistent homology} of the filtration $f_{\leq \bullet}$.

As we have mentioned, the whole process is 1-Lipschitz for appropriate metrics \cite{MR3333456}. The metric of choice to start with on the space of $\R$-valued functions on $X$ is the $L^{\infty}$ distance. The metric of choice on the space of persistence diagrams is the \emph{bottleneck distance}, which measures...

One strategy to prove this is to also define the so-called \emph{interleaving distance} for filtrations and persistence modules \cite{MR2279866}, which measures how much one needs to shift two such objects along the index set $\R$ to obtain an isomorphism between them, and show that each individual step in the pipeline is 1-Lipschitz. With this approach, by far the most difficult step is showing that passing from persistence moduels to persistence diagrams is stable, a result which is known as the Algebraic Stability Theorem  \cite{10.1145/1542362.1542407,Chazal.2016a,MR3348168}.

>>>>

In this section we present an overview of the theory of persistence as it is used in geometric contexts, through filtrations by sublevel sets of real-valued functions.
For a detailed exposition we refer to \cite{polterovich2020topological} and \cite{Chazal.2016a, MR3408277}.

The ``pipeline'' of topological persistence traverses through geometry, algebra and discrete mathematics as follows:
Given a space $X$ filtered by the sublevel sets of a function $f$, the application of some homology theory in degree $n$ with coefficients in a field,  or, more generally, a functor from topological spaces to vector spaces, produces a \textit{persistence module}, an algebraic object equipped with a structure theory leading in favorable cases to a powerful discrete invariant called \textit{persistence diagram}.

These invariants are key to applications.
For example, in the next section we will see how they lead to a generalization of the classical Morse inequalities.
Much of their usefulness comes from a remarkable geometric fact known as stability, stating that the map from functions on $X$ with the supremum norm to the set of all persistence diagrams is 1-Lipschitz with respect to a certain natural metric, called the \emph{bottleneck distance}.

To explain the meaning of the above terms in detail, we consider a space $X$ and a function $f \colon X \to \R$.
Unless noted otherwise, the functions we consider need not be continuous.
We pass to filtered spaces by considering the \textit{sublevel set filtration $f_{\leq \bullet}$ of $X$ induced by $f$}, which is defined by
\begin{equation*}
f_{\leq t} = f^{-1}(-\infty, t].
\end{equation*}
We say that this filtration is \textit{compact} if $X$ is a locally compact Hausdorff space and all sublevel sets are compact.

For the next step in the persistence pipeline, we will need a coherent assignment of a vector space to any topological space, more precisely, a functor from $\Top$ to $\Vect$.
The key examples are provided by \emph{homology theories}, which for now are only assumed to be $\Z$-graded families $\H = (\H_d)_{d \in \Z}$ of \emph{homotopy invariant functors}, meaning that they assign the same morphism to homotopic maps.
Of particular importance to us are \v{C}ech homology \cite[Section IX--X]{MR0050886} with coefficients in some fixed field $\F$, and homology theories in the sense of Eilenberg--Steenrod \cite[Section I]{MR0050886}, such as singular homology, again with coefficients in some field $\F$ \cite{Eilenberg.1944}.

By definition, applying to a filtered space $\{X_t\}_{t \in \R}$ a functor from $\Top$ to $\Vect$ yields for every $t \in \R$ a vector space~$M_t$ and for any pair~$s, t \in \R$ with~$s \leq t$ a linear map~$M_{s,t} \colon M_s \to M_t$ such that $M_{t,t}$ is the identity and the composition $M_{s,t} \circ M_{r,s}$ is equal to $M_{r,t}$ for any triple $r \leq s \leq t$.
In other words, we obtain a functor from the real numbers, considered as a poset category, to the category of vector spaces.
Such functors will be called \emph{persistence modules}.
A morphism of persistence modules $\varphi \colon M \to N$ is a natural transformation, i.e., an assignment of a linear map $\varphi_t \colon M_t \to N_t$ for every $t \in \R$ making the diagram
\begin{equation*}
    \begin{tikzcd}
    M_{s} \arrow[r, "M_{s,t}"] \arrow[d] & M_{t} \arrow[d] \\
    N_{s} \arrow[r, "N_{s,t}"] & N_{t}
    \end{tikzcd}
\end{equation*}
commute for all pairs $s \leq t$.

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for an interval $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
    C(I)_t =
    \begin{cases}
        \mathbb{F} & \text{if } t \in I, \\
        0          & \text{otherwise},
    \end{cases}
    \qquad
    \qquad
    C(I)_{s, t} =
    \begin{cases}
        \operatorname{id}_{\mathbb{F}} & \text{if } s, t \in I,\\
        0 & \text{otherwise}.
    \end{cases}    
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
% \[
% \bigoplus_{\lambda \in \Lambda} C(I_{\lambda}).
% \]
The multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ associated to a barcode module is known as its \textit{barcode}. By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$. Thus, if a persistence module $M$ is isomorphic to a barcode module, the associated barcode is a complete invariant of $M$, and the (non-unique) isomorphism to the barcode module is called a \emph{barcode decomposition}.

We want to use the passage from persistence modules to invariants obtained from decompositions as the last step in our pipeline, so we have to understand which persistence modules admit barcode decompositions.
The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's theorem \cite{Crawley-Boevey.2015}. It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional ($\PFD$)} persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.

Unfortunately, the $\PFD$ condition is too restrictive for our purposes. In particular, it is not necessarily satisfied in the historical setup of Morse and Tompkins' work on minimal surfaces. In this context, the persistence modules $M$ only have the slightly weaker property of being \emph{q-tame}, which is defined by the rank of the maps $M_{s,t} \colon M_s \to M_t$ being finite for all $s < t$  \cite{Chazal.2016a}. 
Not every q-tame persistence module admits a barcode decomposition in the above sense, as exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$. Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}. We briefly recall the approach from \citet{Chazal.2016b}.

A persistence module $K$ is called \emph{ephemeral} if the maps $K_{s,t} \colon K_s \to K_t$ are $0$ for all $s < t$.
The \emph{radical} $\rad M$ of a persistence module $M$ is defined as the unique minimal submodule of $M$ such that the cokernel of the inclusion $\rad M \to M$ is an ephemeral persistence module.
Specifically, $(\rad M)_t = \sum_{s<t}\im M_{s,t}$.
If $M$ is q-tame, then its radical admits a barcode decomposition \cite[Corollary~3.6]{Chazal.2016b},
with the associated barcode describing the isomorphism type of~$M$ ``up to ephemerals".
This can be formalized by constructing the \emph{observable category of persistence modules}, which is equivalent to the quotient of the category of persistence modules by the full sub-category of ephemeral persistence modules.
Intuitively, one may think of the observable category as forgetting all information in persistence modules that does not persist over a non-zero amount of time.
The barcode of the radical of a q-tame persistence module $M$ is then a complete invariant of $M$ in the observable category.

@@

We have thus seen how to obtain a persistence diagram from a real-valued function $f$ by applying a functor $\H \colon \Top \to \Vect$ to its sublevel set filtration and considering the persistence diagram of the resulting persistence module $\H(f_{\leq \bullet})$, which is well-defined provided that $\H(f_{\leq \bullet})$ is q-tame. If this is the case, we will call the function \emph{q-tame} with respect to the functor $\H$. If $\H$ is some kind of homology theory, we will call $\H(f_{\leq \bullet})$ the \emph{persistent homology} of the filtration $f_{\leq \bullet}$.

As mentioned earlier, the passage from functions to persistence diagrams is 1-Lipschitz for appropriate metrics \cite{MR3333456}. The metric of choice to consider on the space of $\R$-valued functions on $X$ is the distance induced by the supremum norm. The metric of choice on the space of persistence diagrams is the \emph{bottleneck distance}, which expresses the distance between persistence diagrams by an optimal matching of the intervals in the corresponding barcodes: two diagrams are within distance $\delta$ if any two matched intervals are within Hausdorff distance $\delta$, and any unmatched interval has length at most $2\delta$.

The most general stability result is shown by considering as an intermediate step the \emph{interleaving distance} for filtrations and persistence modules \cite{MR2279866}, which measures how far two $\R$-indexed diagrams $M,N$ are from being isomorphic.
A $\delta$-interleaving consists of a pair of natural transformations $(M_t \to N_{t+\delta})_t, (N_t \to M_{t+\delta})_t$ from one diagram to a shifted version of the other and vice versa, which compose to the internal structure maps $(M_t \to M_{t+2\delta})_t, (N_t \to N_{t+2\delta})_t$. Clearly, the case $\delta=0$ describes an isomorphism, and the infimum $\delta$ admitting a $\delta$-interleaving is the interleaving distance between the two diagrams.
As it turns out, the stability of persistence barcodes can then be described as a sequence of 1-Lipschitz transformations: from functions (with the supremum norm) to filtrations by sublevel sets (with the interleaving distance on diagrams of topological spaces), to persistent homology (with interleaving distance on persistence modules), and finally to persistence diagrams (with the bottleneck distance).
In this approach, by far the most difficult step is showing that passing from persistence modules to persistence diagrams is stable, a result which is known as the Algebraic Stability Theorem  \cite{10.1145/1542362.1542407,Chazal.2016a,MR3333456}.

### 2021-07-15 surfaces.tex
Before meeting on Friday mid July

They infer the Morse inequalities for homotopicaly critical points by relating these points to cap numbers and then applying Morse inequalities for cap numbers as we have considered them in \cref{s:inequalities} and as Morse considered them in \cite{Morse.1940}.

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, to obtain q-tameness Morse assumed slightly varying forms of local-connectivity on the resulting sublevel set filtrations.
In particular, Morse and Tompkins use in their applications to minimal surface theory the following condition:
\begin{displaycquote}[p.431]{Morse.1940}
	The space $M$ is said to be \textit{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Here, $c = F(p)$.
They then state that as a consequence of these assumptions the persistent \v{C}ech homology of this sublevel set filtration is q-tame.
In the original it reads:
\begin{displaycquote}[Theorem 6.3, p.432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, this argument is not in general sound as we will show next.

\begin{defi}
The sublevel set filtration of a function $f \colon X \to \R$ is said to be $\LC$ 
if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion
\begin{equation*}
f_{\leq s} \cap U \to f_{\leq t} \cap V
\end{equation*}
is homtopic to a constant map.
\end{defi}

Clearly, if the homology $\H$ takes trivial values on one point spaces, being $\LC$ implies being $\HLC$. Moreover, if the sublevel set filtration of a function $F \colon M \to \R$ on a metric space $M$ is $\LC$, then $M$ is locally $F$-connected in the sense of Morse's definition above.

We will now show that neither our $\HLC$ condition nor Morse's condition above are sufficient in general to imply q-tameness for compact sublevel set filtrations induced by non-continuous functions by producing an $\LC$ filtration which is not q-tame. In particular, this invalidates Morse's claim we have quoted above.

To do this, we consider the \textit{$d$-dimensional Hawaiian earring}
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1} \ \middle | \ \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}.
\end{equation*}

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale = 76]
	\draw[thick] (-.1,0) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05); % remove for all circles
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
	\label{fig:earrings}
\end{figure}

@@

Both singular and \v{C}ech homology of the Hawaiian earring are infinite dimensional in some degree, so they satisfy the assumptions of \cref{thm:counterexample}. The singular case is treated in \cite{Barratt.1962}, and for \v{C}ech homology, one can use the fact that it commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6.\@ and X.3.1.]{MR0050886}:
	Define 
	\begin{align*}
	\HE_k &= \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle|\  \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\
	& \cup \bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle|\  \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right \},
	\end{align*}
	i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
	We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \mathbb{F}) = \lim_{k} \CH_{d}(\HE_{k}; \mathbb{F})$.
	Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
	\begin{equation*}
	\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
	\end{equation*}
	which is infinite-dimensional over $\mathbb{F}$.

@@

The gap in the argument of Morse and Tompkins that we have highlighted can be fixed by applying \cref{t:strong local connectedness implies q-tameness} because the proof given in \cite[p.464]{Morse.1939} for local connectivity of $(\Omega_g, A_g)$ actually establishes that the sublevel set filtration of $A_g$ is \emph{strongly} $\LC$, which can be defined analogously to our strong $\HLC$ property. This implies that the filtration is also $\HLC$ for any homology theory that takes trivial values on one point spaces, so in particular for \v{C}ech homology with field coefficients, which is what Morse and Tompkins use.
Therefore, from \cref{t:strong local connectedness implies q-tameness} we can conclude that $(\Omega_g, A_g)$ is q-tame as claimed.
This implies that Morse inequalities hold, and hence so does the unstable minimal surfaces theorem of Morse--Tompkins.

>>>>

They infer Morse inequalities for homotopically critical points by relating these points to cap numbers and then applying the Morse inequalities for cap numbers as we have considered them in \cref{s:inequalities} and as Morse considered them in \cite{Morse.1940}.

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, to obtain q-tameness Morse assumed slightly varying forms of local-connectivity on the resulting sublevel set filtrations.
In particular, Morse and Tompkins use in their applications to minimal surface theory the following condition:
\begin{displaycquote}[p.431]{Morse.1940}
	The space $M$ is said to be \textit{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Here, $c = F(p)$.
Using similar language to the one used in \cref{s:connectivity}, this property is equivalent to the following notion applicable to general topological spaces.

\begin{defi}
    The sublevel set filtration of a function $F \colon M \to \R$ is said to be $\piLC$	if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion
	\begin{equation*}
	f_{\leq s} \cap U \to f_{\leq t} \cap V
	\end{equation*}
	induces the zero map on homotopy groups.
\end{defi}

It is then stated that as a consequence of local $F$-connectivity the persistent \v{C}ech homology of this sublevel set filtration is q-tame.
In the original it reads:

\begin{displaycquote}[Theorem 6.3, p.432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, this argument is not in general sound as we will show next.

\begin{defi}
    The sublevel set filtration of a function $f \colon X \to \R$ is said to be $\LC$ if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with  $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion
    \begin{equation*}
    f_{\leq s} \cap U \to f_{\leq t} \cap V
    \end{equation*}
    is homotopic to a constant map.
\end{defi}

Clearly, being $\LC$ implies being $\piLC$ and, if the homology $\H$ takes finite dimensional values on one point spaces, also $\HLC$.
% Moreover, if the sublevel set filtration of a function $F \colon M \to \R$ on a metric space $M$ is $\LC$, then $M$ is locally $F$-connected in the sense of Morse's definition above.
We will now show that not even $\LC$ is sufficient to ensure the q-tameness of compact sublevel set filtrations.
% neither our $\HLC$ condition nor Morse's condition above are sufficient in general to imply q-tameness for compact sublevel set filtrations induced by non-continuous functions by producing an $\LC$ filtration which is not q-tame.
In particular, our construction will be used to invalidate Morse's claim quoted above.

Consider the \textit{$d$-dimensional Hawaiian earring}
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1} \ \middle | \ \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}.
\end{equation*}

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale = 76]
	\draw[thick] (-.1,-.001) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05); % remove for all circles
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
	\label{fig:earrings}
\end{figure}

@@

Both singular and \v{C}ech homology of the Hawaiian earring are infinite dimensional in some degree, so they satisfy the assumptions of \cref{thm:counterexample}. The singular case is treated in \cite{Barratt.1962}, and for \v{C}ech homology, one can use the fact that it commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6.\@ and X.3.1.]{MR0050886}:
	Define 
	\begin{align*}
	\HE_k &= \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle|\  \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\
	& \cup \bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle|\  \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right \},
	\end{align*}
	i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
	We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \mathbb{F}) = \lim_{k} \CH_{d}(\HE_{k}; \mathbb{F})$, where $\CH$ denotes \v{C}ech homology.
	Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
	\begin{equation*}
	\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \mathbb{F})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
	\end{equation*}
	which is infinite-dimensional over $\mathbb{F}$.

@@

The gap in the argument of Morse and Tompkins that we have highlighted can be fixed by applying \cref{t:strong local connectedness implies q-tameness} because the proof given in \cite[p.464]{Morse.1939} for local connectivity of $(\Omega_g, A_g)$ actually establishes that the sublevel set filtration of $A_g$ is \emph{strongly} $\LC$,  which can be defined analogously to our strong $\HLC$ property.
Explicitly, the sublevel set filtration defined by a function $f \colon X \to \R$ is said to be \emph{strongly $\LC$} if for any $x \in X$, any neighborhood $V$ of $x$ and any pair of indices $f(x) < s < t$ there is a neighborhood $U \subseteq V$ of $x$ such that the map $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to a constant map.
Clearly, the filtration being strongly $\LC$ implies that the filtration is also strongly $\HLC$ 
%for any homology theory that takes trivial values on one point spaces, so in particular
for \v{C}ech homology with field coefficients, which is what Morse and Tompkins use.
Therefore, from \cref{t:strong local connectedness implies q-tameness} we can conclude that $(\Omega_g, A_g)$ is q-tame as claimed.
This implies that Morse inequalities hold, and hence so does the Unstable Minimal Surfaces Theorem by Morse and Tompkins.

We also mention that Morse introduced another condition that he also called local $F$-connectivity three years earlier. 
It roughly corresponds to being \emph{strongly} $\piLC$ with a certain added uniformity property.
In the original it reads:
\begin{displaycquote}[p.421-422]{Morse.1937}
	The space $M$ will be said to be locally $F$-connected for the order $n$ if corresponding to $n$, an arbitrary point $p$ on $M$, and an arbitrary positive constant $e$, there exists a positive constant $\delta$ with the following property. For $c \geq F(p)$ any singular $n$-sphere on $F \leq c$ (the continuous image on $F \leq c$ of an ordinary $n$-sphere) on the $\delta$-neighborhood $p_{\delta}$ of $p$ is the boundary of a singular $(n + 1)$-cell on $F \leq c + e$ and on $p_e$.
\end{displaycquote}
Morse also claims in the given reference that this condition is sufficient for q-tameness, but without providing proof. 
We expect this statement to be true, but do not investigate it further.
